---
title: "Hands-on Exercise 4.1: Geopraphically Weighted Regression"
author: "NeoYX"
date: '4 Dec 2023'
date-modified: "`r Sys.Date()`"
editor: visual
execute: 
  freeze: auto
  warning: false
  #echo: false
  #message: false
format: 
  html:
    code-fold: false
    code-overflow: scroll
    code-summary: "Show the code"
    code-line-numbers: true
---

## **13.1 Overview**

**Geographically weighted regression (GWR)** is a spatial statistical technique that takes non-stationary variables into consideration (e.g., climate; demographic factors; physical environment characteristics) and models the local relationships between these independent variables and an outcome of interest (also known as dependent variable). In this hands-on exercise, you will learn how to build [hedonic pricing](https://www.investopedia.com/terms/h/hedonicpricing.asp) models by using GWR methods. The dependent variable is the resale prices of condominium in 2015. The independent variables are divided into either structural and locational.

## **13.2 The Data**

Two data sets will be used in this model building exercise, they are:

-   URA Master Plan subzone boundary in shapefile format (i.e. *MP14_SUBZONE_WEB_PL*)

-   condo_resale_2015 in csv format (i.e. *condo_resale_2015.csv*)

## **13.3 Getting Started**

Before we get started, it is important for us to install the necessary R packages into R and launch these R packages into R environment.

The R packages needed for this exercise are as follows:

-   R package for building OLS and performing diagnostics tests

    -   [**olsrr**](https://olsrr.rsquaredacademy.com/)

-   R package for calibrating geographical weighted family of models

    -   [**GWmodel**](https://cran.r-project.org/web/packages/GWmodel/)

-   R package for multivariate data visualisation and analysis

    -   [**corrplot**](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html)

-   [ggpubr](https://rpkgs.datanovia.com/ggpubr/) for some easy-to-use functions for creating and customizing 'ggplot2'- based publication ready plots.

-   Spatial data handling

    -   **sf** for importing, integrating, processing and transforming geospatial data.

-   Attribute data handling

    -   **tidyverse**, especially **readr**, **ggplot2** and **dplyr**

-   Choropleth mapping

    -   **tmap** for creating thematic maps

The code chunks below installs and launches these R packages into R environment.

```{r}
#| eval: false
#| echo: false

pacman::p_load(olsrr, corrplot, ggpubr, sf, spdep, GWmodel, tmap, tidyverse, gtsummary, DT, knitr)
```

```{r}

#| eval: false
#| echo: false
#| fig-width: 14
#| fig-asp: 0.68
#| code-fold: True
```

## Summaries
